A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-08-29 paper/paper.discussion.tex
SM edits/comments discussion

In this paper our goal was to develop a rather simple, yet flexible, integer linear programming formulation to generate schedules for clinical departments at hospitals. The difficulty in applying such an approach to this task lies in the fact that ILP is an NP-hard problem, and as such, the time to find an optimal solution grows exponentially as we increase the size of the problem. As a result, the vast majority of approaches taken to create schedules for similar scenarios tend to use heuristics in order to find an approximately optimal solution in a shorter time. \\

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. While one of our goals was to keep a minimal set of constraints to ensure that the ILP formulation is usable in practice, it leaves room to adapt it to various clinical departments at different hospitals. The flexibility of our ILP allows changing the number of services provided in a division, the length of a work block, clinicians' preference for block to weekend adjacency as well as clinicians' requests for time off. \\

We developed a tool that implements the ILP formulation and generates an optimal schedule based on provided input data. It provides a simple and easy-to-use user interface that can be used by any hospital staff to generate schedules well in advance. We compared the optimal schedule generated with the help of the tool to the manually-created schedules at St. Michael's Hospital. We found that the ILP formulation was always able to find an optimal schedule satisfying all required hard constraints, unlike the manual schedule, that often did not take all constraints into account. Moreover, due to the multi-goal objective function in the ILP, the tool was able to maximize and fulfill the majority of clinician preferences and requests, more so than the manually-created schedule. These observations reinforce the idea that it is vital to employ automated tools when generating schedules in hospital departments, to reduce human error, balance the work-load of clinicians and improve the service provided to patients. \\

We analyzed the performance of our formulation on simulated data meant to resemble real-life clinician departments at different hospitals. Our results show that increasing the number of requests per clinician does not affect the runtime of the algorithm, highlighting the flexibility of the tool in incorporating clinician preferences. Further, we saw that the algorithm can accommodate increasing the time-horizon up to four years with little impact on runtime, which can be applied to departments that generate schedules far in advance. The most notable impact on the runtime was noticed when trying to increase the number of services offered in a single department. While such cases are unlikely to be encountered in the real-world, as most departments tend to focus on a single service [ref \ref{???}], the runtime issues can be mitigated by relaxing certain constraints, which can then be readjusted manually from the generated schedule. Overall, it is very encouraging to observe that an ILP formulation can be applied to schedule clinicians in practice, by providing optimal schedules rather than having to utilize heuristics. \\

In future work, we can look first and foremost into improving the runtime of the tool by modifying constraints and objectives, so that it can be used in larger departments offering multiple services for their patients. On top of that, we can introduce new objectives that incorporate additional clinician preferences and allow for better distribution of work-load. However, since our ultimate goal is to maintain a ILP formulation, care must be taken not to overly complicate the model and avoid running into additional time complexity issues.

>>>>

In this paper, we present a simple, yet flexible, integer linear programming formulation to generate schedules for clinical departments at hospitals. 
The challenge in applying such an approach to this task lies in the fact that ILP is an NP-hard problem. %specify what 'such an approach' and 'this task' are... the first part of the sentence was a bit hard to follow. Define what you mean by NP-hard problem and cite --> i.e. NP-hard = the following sentence about how time to optimal solution grows exponentially? if yes, then instead of 'As such' - say 'That is,...or "An NP-hard problem means that..."
As such, the time to find an optimal solution grows exponentially as we increase the size of the problem. As a result, the vast majority of approaches taken to create schedules for similar scenarios tend to use heuristics in order to find an approximately optimal solution in a shorter time. \\  %cite/reference statement (i.e. the 'vast majority'...) I can't remember if the introduction details the heuristics --> but if yes, then great. if no, then include that review of the prior literature here in discussion. I think as you said - the journals we are submitting to seem to have more of those details in the introduction.

%double-check all abbreviations have been defined. I would prefer limiting abbreviations to no more than 3 at most if possible. We are ok for word count in this paper :).

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. Although we restricted our use of the formulation to a minimal set of constraints for the particular needs of the case study (St. Michael's Hospital Division of Infectious Diseases) - our formulation can be adapted to various clinical departments at different hospitals. The flexibility of our ILP allows changing the number of services provided in a division, the length of a work block, clinicians' preference for block to weekend adjacency as well as clinicians' requests for time off. \\

When comparing the optimal schedule generated by tool to the manually-created schedules at St. Michael's Hospital, we found that the ILP formulation was always able to find an optimal schedule satisfying all required hard constraints, unlike the manual schedule, that often did not take all constraints into account. Moreover, due to the multi-goal objective function in the ILP, the tool was able to maximize and fulfill the majority of clinician preferences and requests, more so than the manually-created schedule. These observations reinforce the benefits of automated tools when generating schedules in hospital departments to balance the work-load of clinicians and improve the service provided to patients. \\  % how the finding compares to the wider literature when automated compared to manual? What does it mean re: "so what" - human error in generating tools, etc. or challenges in heuristics used manually by people? include citations this last sentence in particular - what dos it reinforce exactly - what are the benefits?

We also found that increasing the number of requests per clinician does not affect the runtime of the algorithm, highlighting the flexibility of the tool in incorporating clinician preferences. Further, we saw that the algorithm can accommodate increasing the time-horizon up to four years with little impact on runtime, which can be applied to departments that generate schedules far in advance. One key limitation we identified was the sensitivity of the runtime to larger numbers of services offered in a single department. Such cases are unlikely to be encountered in the real-world because most clinical departments tend to provide a single service or at most two services by the same roster of clinicians [ref \ref{???}]. One solution to mitigate the runtime issues generated by a larger number of services would be relaxing certain constraints, followed by manual readjustment from the generated schedule. %what do we mean by certain constraints? specify the other constraints?
Thus, next steps include expanding the generalizability of the tool beyond smaller clinical departments to larger departments within and outside of health-care - especially those that provide multiple services i parallel for patients and other clients. Overall, our sensitivity analyses using simulated data provide reassurance that the ILP formulation can be applied to schedule clinicians across real-world variability between clinical departments. \\ %SM - will look for reference and we can ask Kevin Gough too.
